---
title: "EpiFunctions"
output: 
  github_document:
    toc: true
always_allow_html: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "35%"
)

library(dplyr)
library(readr)
library(knitr)
library(kableExtra)
library(EpiFunctions)
library(LTASR)
library(ggplot2)
library(dplyr)
library(tidyr)
per_form <- tibble(Variable = c('id',
'sex',
'race',
'dob',
'pybegin',
'dlo',
'...'
),
Description = c('Unique identifier for each person',
'Sex of person ("M" = male / "F" = female)',
'Race of person ("W" = white / "N" = nonwhite)',
'date of birth',
'date to begin follow-up',
'date last observed. Minimum of end of study, date of death, date lost to follow-up',
'Additional variable(s) time-fixed strata and variables defining death/incidence outcome'
),
Format = c('',
'character',
'character',
'date',
'date',
'date',
'...'
)
)

hist_form <- tibble(Variable = c('id',
'begin_dt',
'end_dt',
'\\<daily exposure variables\\>'
),
Description = c('Unique identifier for each person',
'Beginning date of exposure period',
'End date of exposure period',
'Exposure variable(s)'
),
Format = c('',
'character',
'character',
'numeric'
)
)
```

<!-- badges: start -->
<!-- badges: end -->

The goal of EpiFunctions is to provide a set of functions to perform survival analysis regressions not found elsewhere. In particular, background stratified Poisson regression, as described in Richardson and Langholz (2012), and matched Cox regression. Both regressions allow for specification of the relative risk (RR) function as the traditional log-linear form: [RR = exp(b * exposure)] as well as a linear form: [RR = (1 + b * exposure)] commonly used in radiation research to calculate Excess Relative Risk (ERR). Additionally, it provides options for the calculation of profile likelihood confidence intervals (vs the traditional Wald-based confidence intervals).  

Additionally, tools are provided to implement each regression in the presence of time-dependent exposures.  

## Installation

You can install the development version of EpiFunctions from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("bertkesj/EpiFunctions")
```

## Data format  
### Person File  

A person file is required to run most analyses where each row represents one person and contains the following information:  

```{r, fig.align='center', echo = FALSE}
kable(per_form) %>%
  kable_styling(c("striped"), 
                 full_width = FALSE)
```

EpiFunctions comes with a 'dummy' person data.frame, named `example_person` for illustration and testing.  

```{r, echo=FALSE}
example_person %>%
  head() %>%
  kable()%>%
  kable_styling(c("striped"), 
                 full_width = FALSE)
```  
Currently in this file, dates are saved as characters, sex is contained in a variable named `gender` and the outcome variable is named `lung_cancer`. These will need to be adjusted before most analyses (See next sections).  

### History File  

Information on time-dependent covariates (aside from age and calendar period) will be contained in a history file. This file contains one row per person per exposure period. An exposure period is a period of time in which all daily values/levels of an exposure variable are assumed to be constant.  

Below are the required variables to be found within the history file:

```{r, fig.align='center', echo = FALSE}
kable(hist_form) %>%
  kable_styling(c("striped"), 
                 full_width = FALSE)
``` 

EpiFunctions comes with a 'dummy' history file also, named `example_history`, for illustration and testing.  

Below are the entries for person id '198' in the `example_history` file:  
```{r, echo=FALSE}
example_history %>%
  filter(id == '198') %>%
  head(21) %>%
  kable()%>%
  kable_styling(c("striped"), 
                 full_width = FALSE)
```   
Again, dates are saved as characters and will need to be converted.  

The above person's history file entry contains 5 exposure periods, one of which was unexposed. Below plots their cumulative exposure (which starts at 0) over time:  
```{r, echo=FALSE, fig.align='center', fig.width=3, fig.height=3}
person <- example_person %>%
  mutate(dob = as.Date(dob),
         pybegin = as.Date(pybegin),
         dlo = as.Date(dlo))
history <- example_history %>%
  mutate(begin_dt = as.Date(begin_dt),
         end_dt = as.Date(end_dt),
         exposure_level = as.numeric(daily_exposure))
pid <- '198'
history %>%
  filter(id ==pid) %>%
  bind_rows(tibble(id=pid, 
                   begin_dt=filter(history, id==pid, row_number()==n())$end_dt + 1,
                   end_dt=filter(person, id==pid)$dlo,
                   exposure_level=0)) %>%
  expand_dates(begin_dt, end_dt) %>%
  filter(!is.na(period)) %>%
  mutate(Date = date,
         `Exposure Level` = cumsum(exposure_level)) %>%
  select(id, Date, `Cumulative Exposure` = `Exposure Level`)%>%
  
  ggplot() +
  geom_line(aes(x=Date,
                y=`Cumulative Exposure` )) +
  theme_bw()

```  
  
Note that cumulative exposure increases by 6.5 units per day and the rate of increase drops to 0.6 units per day towards the end of their exposure history.  

**NOTE:** Any gaps within the history file and the follow-up times (for example, the period between the last exposure period within the history file through the end of follow-up) are assumed to be 0. That is, cumulative exposure does not change during these periods.

## Cox Regression  
To perform Cox regression, first risk-sets must be assembled. The function `gt_rs()` will read in the person file (which is required to contain id, race, sex, dob, pybegin, dlo) and also requires a variable called `case` that takes a TRUE/FALSE value identifying the cases of interest. 

```{r}
library(EpiFunctions)
risk_sets <- example_person %>%
  mutate(dob = as.Date(dob),
         pybegin = as.Date(pybegin),
         dlo = as.Date(dlo),
         
         sex = gender, #renaming gender to sex
         
         case = (lung_cancer == 'TRUE')) %>%
  gt_rs()
```  
```{r, fig.align='center', echo = FALSE}
risk_sets %>%
  head(10) %>%
  kable() %>%
  kable_styling(c("striped"), 
                 full_width = FALSE)
``` 
The above risk sets use age as the time-scale. The `case_id` indicates the index case and `cut_dt` indicates the date the given person (`id`) attains the age of the index case. This date is used when evaluating time-dependent covariates.   

**NOTE:** The risk-sets are 'matched' on sex, race and birth-date (+/- 5 years) by default and therefore, these confounders do not need to be explicitly controlled for in any regression. 

Once the risk-sets are formed, they can be merged to the person file to add additional confounders of interest. To add time-dependent exposures found in the history file, the helper function `cum_exp()` function is useful.  

Below merges the history file with the risk_sets and calculates cumulative exposure up to the `cut_dt` with a 0 and 10 year lag:

```{r}
risk_sets <- example_history %>%
  mutate(begin_dt = as.Date(begin_dt),
         end_dt= as.Date(end_dt),
         
         daily_exposure = as.numeric(daily_exposure)) %>%
  right_join(risk_sets,
             by='id',
             relationship = 'many-to-many') %>%
  mutate(cumulative_exposure_lag0 = cum_exp(daily_exposure, 
                                             begin_dt, end_dt, cut_dt, 0),
         cumulative_exposure_lag10 = cum_exp(daily_exposure, 
                                             begin_dt, end_dt, cut_dt, 10)) %>%
  group_by(case_id, id, case, cut_dt) %>%
  dplyr::summarize(cumulative_exposure_lag0 = sum(cumulative_exposure_lag0),
                   cumulative_exposure_lag10 = sum(cumulative_exposure_lag10),
                   .groups='drop')
```
```{r, fig.align='center', echo = FALSE}
risk_sets %>%
  arrange(as.numeric(case_id), as.numeric(id)) %>%
  head(10) %>%
  kable() %>%
  kable_styling(c("striped"), 
                 full_width = FALSE)
```  

To perform Cox regression on these risk-sets, the `coxphreg()` function is used. Below calculates log(HR) for a 10 year lagged exposure. Note, due to the high value of exposures, these values are scaled and the below represent log(HR) per 100,000 unit increase in dose, matched on age, race, sex, birth-cohort:  

```{r}
coxphreg(risk_sets,
         loglin = ~ scale(cumulative_exposure_lag10, scale = 100000))
```  
```{r, echo=FALSE}
cox_out <- coxphreg(risk_sets,
         loglin = ~ scale(cumulative_exposure_lag10, scale = 100000)) %>%
  `$`(output)

```  

Note that this can also be accomplished through the use of the `survival` package, where matching on race, sex and birth cohort being the equivalent of a stratified Cox regression on these variables. However, the `survival` package does not allow for specification of the hazard ratio with a linear term (i.e. ERR calculation).  

Additionally, `EpiFunctions` matches on birth dates within 5 years of the index case (known as caliper matching), whereas stratified Cox regression with the `survival` package defines fixed categories. See "Modern Epidemiology" (Ahlbom et al., 2021) for further discussion of this difference. Below nearly replicates the above analysis (with birth cohort instead controlled for through defining fixed 10-year categories):  

```{r}
library(lubridate)
library(survival)
person <- person %>%
  mutate(age_begin = difftime(pybegin, dob, units = 'days') %>%
           as.numeric() %>%
           `/`(365.25),
         age_end = difftime(dlo, dob, units = 'days') %>%
           as.numeric() %>%
           `/`(365.25),
         birth_cat = 10*floor(year(dob)/10))
event_times <- person %>%
  filter(lung_cancer == 'TRUE') %>%
  `$`(age_end) %>%
  sort()
person_time <- person %>%
  mutate(event_times = list(event_times)) %>%
  unnest(event_times) %>%
  filter(age_begin <= event_times & event_times <= age_end) %>%
  group_by(id) %>%
  mutate(cut_dt = dob + 365.25*event_times,
         beg = if_else(row_number() == 1, 
                       age_begin, 
                       lag(event_times)),
         end = event_times,
         case = if_else(row_number() == n(),
                        lung_cancer == 'TRUE',
                        FALSE)) %>%
  select(id, race, gender, birth_cat, beg, end, case, cut_dt)
person_time <- example_history %>%
  mutate(begin_dt = as.Date(begin_dt),
         end_dt= as.Date(end_dt),
         
         daily_exposure = as.numeric(daily_exposure)) %>%
  right_join(person_time,
             by='id',
             relationship = 'many-to-many') %>%
  mutate(cumulative_exposure_lag10 = cum_exp(daily_exposure, 
                                             begin_dt, end_dt, cut_dt, 
                                             10)) %>%
  group_by(across(colnames(person_time))) %>%
  summarize(cumulative_exposure_lag10 = sum(cumulative_exposure_lag10),
            .groups='drop')
coxph(Surv(beg, end, case) ~ strata(race, gender, birth_cat) +
        scale(cumulative_exposure_lag10, scale = 100000),
        data = person_time,
        ties='breslow')
        
```  
```{r, echo=FALSE}
cox_mod <- coxph(Surv(beg, end, case) ~ strata(race, gender, birth_cat) +
        scale(cumulative_exposure_lag10, scale = 100000),
        data = person_time,
        ties='breslow')
cox_surv <- tibble(est = cox_mod$coefficients,
                   se = sqrt(cox_mod$var))
```
Another drawback of this method is computational. The `survival` package requires expanding the person file to each observed event-time. This is similar to the formation of risk-sets, however, matching on additional confounders (gender, race, birth date) significantly reduces the size of this file and speeds up computation time.  

## Background Stratified Poisson Regression  
Poisson regression is another method for analyzing survival data. A first step of performing poisson regression is to stratify person-time information. `EpiFunctions` provides a function that will stratify a person data.frame into age and calendar period strata (as well as other time-fixed user-defined strata). This function is `get_table_rapid()` and runs very fast by taking advantage of the regularity in which age and calendar period change over time (see Macaluso, 1992).  

To handle additional time-dependent covariates found in a history data.frame, the package `LTASR` can be used. This is accomplished with `get_table_history_est()`. Details of these functions can be found in the `LTASR` documentation (https://CRAN.R-project.org/package=LTASR), namely the [History File Stratification](https://cran.r-project.org/web/packages/LTASR/vignettes/HistFile.html) vignette.  

Below will create a stratified table based on information in the `example_person` and `example_history` data.frames (as detailed in the previously mentioned vignette):  

```{r, message=FALSE}
library(LTASR) 
# Load data from EpiFunctions package
data(example_person,
               package='EpiFunctions')
data(example_history,
               package='EpiFunctions')

# Format person and history files
person <- example_person %>%
  mutate(dob = as.Date(dob),
         pybegin = as.Date(pybegin),
         dlo = as.Date(dlo),
         
         vs = if_else(lung_cancer == 'TRUE',
                      'D',
                      'A'),
         rev = if_else(lung_cancer == 'TRUE',
                      10,
                      NA),
         code = if_else(lung_cancer == 'TRUE',
                      'C34',
                      NA))
history <- example_history %>%
  mutate(begin_dt = as.Date(begin_dt),
         end_dt = as.Date(end_dt),
         
         daily_exposure = as.numeric(daily_exposure))


# Define Exposure Stratification Cutpoints
exp <- exp_strata(var = 'daily_exposure',
                  cutpt = c(-Inf, 0, 5, 10, 25, 50, 100, 250, 500, 1000, 
                            5000, 10000, 20000, Inf),
                  lag = 10)

py_table <- get_table_history_est(person,
                                  us_119ucod_19602021,
                                  history,
                                  exps = list(exp))


```  

```{r, echo=FALSE}
library(stringr)
py_table %>% 
  ungroup() %>%
  mutate(across(c(ageCat, CPCat, daily_exposureCat),
         ~ str_remove_all(., '\\[|\\)|\\]|\\('))) %>%
  rename(o16 = `_o16`) %>%
  
  head(5) %>%
  kable() %>%
  kable_styling(c("striped"), 
                 full_width = FALSE)
```   
  
This stratified person-time table has one outcome (saved as variable `_o16`) from the person file, which was recoded as lung cancer ICD-10 code C34 and mapped to LTAS minor 16. It has person days (`pdays`) and outcomes stratified into 5-year age (`ageCat`), 5-year calendar period (`CPCat`), gender, race and exposure (`daily_exposureCat`) categories. The person-time weighted mean exposure (`daily_exposure`) for each strata are also included and can be treated as a continuous variable in future regressions.  

The function `poi_back` will performed background stratified poisson regression on this data. It reads in a stratified table, which is required to include a variable named `pdays` indicating the amount of person-time (outputted as number of days from `get_table_history_est()`) for a given strata. The user can then define which confounders to use as stratification in the regression (below are age, calendar period, gender and race wrapped in `vars()`) and the variable indicating the outcome counts (below `_o16`): 
```{r}
poi_back(py_table,
         quo(`_o16`),
         ss=vars(ageCat, CPCat, gender, race),
         loglin = ~ scale(daily_exposure, scale = 100000))
```  
```{r, echo=FALSE}
poi_out <- poi_back(py_table,
         quo(`_o16`),
         ss=vars(ageCat, CPCat, gender, race),
         loglin = ~ scale(daily_exposure, scale = 100000)) %>%
  `$`(output)
```

Note this could also be accomplished through R's `glm()` function:  
```{r}
model <- glm(`_o16` ~ offset(log(pdays)) +
               factor(paste0(ageCat, CPCat, gender, race)) +
               scale(daily_exposure, scale = 100000),
             data = py_table,
             family = poisson(link = 'log'))

library(stringr)
summary(model)$coefficients %>%
  as_tibble(rownames = 'var') %>%
  filter(str_count(var, 'daily_exposure') > 0)
```  
```{r, echo=FALSE}
poi_glm <- summary(model)$coefficients %>%
  as_tibble(rownames = 'var') %>%
  filter(str_count(var, 'daily_exposure') > 0) %>%
  select(est = Estimate, se = `Std. Error`)
```

The disadvantage of this is that all stratification variables are estimated explicitly. In this case, the poisson regression estimates `r nrow(summary(model)$coefficients)` covariates (1 exposure and `r nrow(summary(model)$coefficients) - 1` strata combinations). The function `poi_back()` treats these `r nrow(summary(model)$coefficients) - 1` covariates as nuisance parameters and do not explicitly estimate them. Therefore, `poi_back()` will typically run faster. 

Additionally, `glm()` does not allow for the linear/ERR specification of the RR function.  

## Final Notes  
It is worth at this point to note that we have fit four different regressions to the example data in this cohort: 2 Cox regressions and 2 poission regressions. They each are attempting to estimate the effect of exposure on the outcome while controlling for age, calendar period / birth date, race and sex.  

Controlling for birth date and age implicitly controls for calendar period since birth date + age will perfectly define the current date. Similarly, controlling for age and birth date implicitly controls for calendar period. Therefore, the above models should largely be equivalent.  

Below plots the outcome from each of the above regressions:  
```{r, echo=FALSE, fig.align='center', fig.width=8, fig.height=6}

bind_rows(mutate(poi_glm, model = 'Poisson glm'),
          mutate(poi_out, model = 'Poisson poi_back'),
          mutate(cox_out, model = 'Cox coxphreg'),
          mutate(cox_surv, model = 'Cox coxph')) %>%
  mutate(model = factor(model, 
                        c('Cox coxphreg', 'Cox coxph',
                          'Poisson poi_back', 'Poisson glm'),
                        c('Cox coxphreg', 'Cox coxph\nSurvival package',
                          'Poisson poi_back', 'Poisson glm\nbase R'))) %>%
  mutate(lower = est - 1.96*se,
         upper = est + 1.96*se) %>%
  ggplot() +
  geom_point(aes(x=model,
                 y=est)) +
  geom_errorbar(aes(x=model,
                    ymin=lower,
                    ymax=upper),
                width = .25) +
  theme_bw() +
  xlab('') +
  ylab('log(RR) per 100,000 units with 95% CIs')
```
  
These estimates are largely equivalent. The differences are due to slight differences in how the covariates are specified. Namely, sex and race are categorical and treated the same in each model. The difference in age and calendar period / birth date. These are summarized below:

```{r, echo=FALSE}
tribble(
  ~ ` `, ~`birth date / calendar period`, ~`age`,
"coxphreg",	"birth date caliper matched within 5 years of case", "exact control through matching",
"coxph",	"birth date as predefined 10-year fixed categories", "exact control through matching",
"poi_back",	"5 year calendar period categories",	"5 yr categories",
"glm",	"5 year calendar period categories",	"5 yr categories",
) %>%
    kable() %>%
  kable_styling(c("striped"), 
                 full_width = FALSE)
```

Generally, since the Cox regressions match on exact age, it provides more precise control as opposed to the categorical specification (5-year categories) used in Poisson regression.  

Additionally, while Cox regression generally imposes the 'proportional hazards' assumption, that same assumption is also imposed in these poisson regressions. In this setting, the proportional hazard assumption assumes that the effect of exposure on risk is the same across ages (i.e. there is no interaction between the time-scale age and exposure). It is a result from how the risk function is defined, which is the same in both types of regression, and does not include an age x exposure interaction term.  

Finally, while the above examples can be replicated through existing packages, the current package allows for two extensions:  
1. defining the risk function as a linear (vs log-linear) function of exposure and  
2. outputting profile likelihood confidence intervals by setting `gcis = TRUE`  


#### References
Richardson, David B., and Bryan Langholz. "Background stratified Poisson regression analysis of cohort data." Radiation and environmental biophysics 51 (2012): 15-22.  

Ahlbom A. Modern Epidemiology, 4th edition. TL Lash, TJ VanderWeele, S Haneuse, KJ Rothman. Wolters Kluwer, 2021. Eur J Epidemiol. 2021 Aug;36(8):767-768.  

Macaluso, Maurizio. "Exact stratification of person-years." Epidemiology 3.5 (1992): 441-448.
